---
title: "Homework 3"
author: "Jared Klug"
output: github_document
---
```{r load_lib}

library(tidyverse)
library(p8105.datasets)

```

## Problem 1

## Problem 2

```{r read_activity_df}

activity_df = read.csv("./data/accel_data.csv") %>% 
  pivot_longer(activity.1:activity.1440,
               names_to = "minute_of_day",
               names_prefix = "activity.",
               values_to = "activity_count") %>% 
  mutate(weekday_or_weekend = ifelse(day == "Saturday" | day == "Sunday", "weekend", "weekday"),
         minute_of_day = as.integer(minute_of_day))

```

The accelerometer data taken from a 63 yr-old male with BMI of 25, with `r nrow(activity_df)` observations and `r ncol(activity_df)` variables, contains the following variables:  

* **week**: `r activity_df %>% pull(week) %>% typeof()` type designating which week the data was recorded from week `r activity_df %>% pull(week) %>% min()` to week `r activity_df %>% pull(week) %>% max()`

* **day_id**: `r activity_df %>% pull(day_id) %>% typeof()` type designating which day of the 5 week period the data was recorded from day `r activity_df %>% pull(day_id) %>% min()` to day `r activity_df %>% pull(day_id) %>% max()`

* **day**: `r activity_df %>% pull(day) %>% typeof()` type designating the day of the week to the corresponding data

* **minute_of_day**: `r activity_df %>% pull(minute_of_day) %>% typeof()` type designating the minute of the day starting at midnight from minute `r activity_df %>% pull(minute_of_day) %>% min()` to minute `r activity_df %>% pull(minute_of_day) %>% max()`

* **activity_count**: `r activity_df %>% pull(activity_count) %>% typeof()` type that measures accelerometer data that can be translated into physcial activity levels. 

* **weekday_or_weekend**: `r activity_df %>% pull(weekday_or_weekend) %>% typeof()` type that says if the corresponding data was taken on a weekday or weekend



## Problem 3