---
title: "Homework 3"
author: "Jared Klug"
output: github_document
---
```{r load_lib}

library(tidyverse)
library(p8105.datasets)
library(ggridges)
library(patchwork)

```

## Problem 1 

```{r load_instacar}

data("instacart")

```

The instacart dataset comes from Instacart, an online grocery shopping service, from 2017. With `r nrow(instacart)` observations and `r ncol(instacart)` variables, instacart contains the following variables:

* **order_id**: Order identifier number (1 - 3421070)

* **product_id**: Product identifier number (1-49688)

* **add_to_cart_order**: Designates order at which product was added to cart by order_id

* **reordered**: Indicates if product has been ordered by user in the past. (1: yes, 0: no)

* **user_id**: Customer identification number

* **eval_set**: Which evaluation set this order belongs in

* **order_number**: Indicates order sequence of order_id by user

* **order_dow**: Indicates day of the week the order was placed (0-6)

* **order_hour_of_day**: Indicates hour of the day the order was placed (0-23)

* **days_since_prior_order**: Days since last order (0-30, if more than 30 days - observation is 30)

* **product_name**: Name of the product

* **aisle_id**: Aisle Identification Number (1-134)

* **department_id**: Department Identification Number (1-21)

* **aisle**: Name of the aisle by aisle_id

* **department**: Name of the department by department_id

* **How many aisles are there, and which aisles are the most items ordered from?**

This dataset contains info on both the user and the products, allowing us to identify trends in either user behavior or product specifications. Across the `r instacart %>% pull(department_id) %>% max()` departments, there are `r instacart %>% pull(aisle_id) %>% max()` aisles and `r instacart %>% pull(product_id) %>% max()` unique products to choose from. 

```{r count_aisle}
instacart %>% 
	count(aisle) %>% 
	arrange(desc(n))
```

The most popular items ordered are fruits and vegetables, followed by dairy, water, and snack products. 

* **Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.** 

```{r plot_aisle_ordered}
instacart %>% 
	count(aisle) %>% 
	filter(n > 10000) %>% 
	mutate(
		aisle = factor(aisle),
		aisle = fct_reorder(aisle, n)
	) %>% 
	ggplot(aes(x = aisle, y = n)) + 
	geom_point() + 
	theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs( 
    title = "Number of Items Ordered in each Aisle",
    x = "Aisle", 
    y = "Number Ordered"
  )
  
```

As pointed out by the count function before, we see the same top aisles to be fruits and veggies, this plot is able to visualize just how much more fruits and vegetables are ordered compared to any other item by aisle. 

* **Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table**

```{r table_popular_in_aisle}

instacart %>% 
	filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>% 
	group_by(aisle) %>% 
	count(product_name) %>% 
	mutate(rank = min_rank(desc(n))) %>% 
	filter(rank < 4) %>% 
	arrange(aisle, rank) %>% 
  select(-rank) %>% 
	knitr::kable()

```

In the baking ingredients, the 3 most popular items are light brown sugar, pure backing soda, and cane sugar. For dog food care, the top ordered items are treats, oragnix dog food, and small dog biscuits. The most popular packaged vegetable/fruit is baby spinach followed by raspberries then blueberries. 

* **Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).**

```{r table_apples_vs_icecream}
instacart %>% 
	filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>% 
	group_by(product_name, order_dow) %>% 
	summarize(mean_hour = mean(order_hour_of_day)) %>% 
	pivot_wider(
		names_from = order_dow,
		values_from = mean_hour
	) %>% 
  knitr::kable()
```

From this table we can see that apples are typically ordered before ice cream, with the exception of the weekday 5. We can assume that earlier in the day people may be more inclined to order healthy foods while later in the day people are more likely to indulge in their sweet-tooth craving. 

## Problem 2

* **Load, tidy, and otherwise wrangle the data. Your final dataset should include all originally observed variables and values; have useful variable names; include a weekday vs weekend variable; and encode data with reasonable variable classes. Describe the resulting dataset (e.g. what variables exist, how many observations, etc).**

```{r read_activity_df}

activity_df = read.csv("./data/accel_data.csv") %>% 
  pivot_longer(activity.1:activity.1440,
               names_to = "minute_of_day",
               names_prefix = "activity.",
               values_to = "activity_count"
  ) %>% 
  mutate(
    weekday_or_weekend = ifelse(day == "Saturday" | day == "Sunday", "weekend", "weekday"),
    minute_of_day = as.integer(minute_of_day)
  ) %>% 
  mutate(
    day = factor(day, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
  )

```

The accelerometer data taken from a 63 yr-old male with BMI of 25, with `r nrow(activity_df)` observations and `r ncol(activity_df)` variables, contains the following variables:  

* **week**: `r activity_df %>% pull(week) %>% class()` type designating which week the data was recorded from week `r activity_df %>% pull(week) %>% min()` to week `r activity_df %>% pull(week) %>% max()`

* **day_id**: `r activity_df %>% pull(day_id) %>% class()` type designating which day of the 5 week period the data was recorded from day `r activity_df %>% pull(day_id) %>% min()` to day `r activity_df %>% pull(day_id) %>% max()`

* **day**: `r activity_df %>% pull(day) %>% class()` type designating the day of the week to the corresponding data

* **minute_of_day**: `r activity_df %>% pull(minute_of_day) %>% class()` type designating the minute of the day starting at midnight from minute `r activity_df %>% pull(minute_of_day) %>% min()` to minute `r activity_df %>% pull(minute_of_day) %>% max()`

* **activity_count**: `r activity_df %>% pull(activity_count) %>% class()` type that measures accelerometer data that can be translated into physcial activity levels. 

* **weekday_or_weekend**: `r activity_df %>% pull(weekday_or_weekend) %>% class()` type that says if the corresponding data was taken on a weekday or weekend

* **Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate accross minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?**

```{r table_activity}

activity_df %>% 
  group_by(week, day) %>% 
  summarize(total_activity = sum(activity_count)) %>% 
  pivot_wider(
    names_from = "day",
    values_from = "total_activity"
  ) %>%
  knitr::kable(digits = 0)
  
```

It is hard to detect any trends in the table without more exploratory analysis. 

* **Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.**

```{r plot_activity}

activity_df %>%
  ggplot(aes(x = minute_of_day, y = activity_count, color = day)) + 
  geom_smooth(se = FALSE) +
  labs(
    title = "24Hr Activity Plot Over 5 Weeks",
    x = "Minute of the Day (min)", 
    y = "Activity Count (arbitrary unit)",
    caption = "Accelerometer data of a 63yr old male with BMI of 25 over 5 weeks"
  )
  
```

There were too many points to plot for a scatterplot to be interpreted. I opted to only keep a smoothed version as it will average the activity counts over the weeks for each day and present a more aesthetic graph. 

From the plot we can see that at the start of the day we see very low activity, assuming that is this particular person sleeping. At around 6 am this person seems to wake up and start their day, indicated by the rise in activity level at around the 250-300 minute mark. This person then remains awake/active steadily throughout the week from minute 375 to 1250. This person seems to be less active during Saturdays compared to the rest of the week. This person also seems to be more active in the later half of the day with the exception of Sunday.  
 
## Problem 3

```{r load_ny_noaa}

data("ny_noaa")

```

This dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. 

Observations made in this dataset are from NY based weather stations. The following variables are included in this dataset:

* **id**: There are `r ny_noaa %>% distinct(id) %>% nrow` distinct weather stations reporting data

* **date**: The dates corresponding to the reported weather data are from `r ny_noaa %>% pull(date) %>% min()` -- `r ny_noaa %>% pull(date) %>% max()`

* **prcp**: Precipitation data reported in tenths of mm are reported. Indicates how much rain on a given day (# NAs: `r ny_noaa %>% pull(prcp) %>% is.na() %>% sum()`)  

* **snow**: Snowfall data reported in mm. Indicates how much snow has fallen on a given day (# NAs: `r ny_noaa %>% pull(snow) %>% is.na() %>% sum()`) 

* **snwd**: Snow depth data reported in mm. Indicates the max depth of snow on the ground on a given day (# NAs: `r ny_noaa %>% pull(snwd) %>% is.na() %>% sum()`) 

* **tmax**: Maximum temperature on a given day reported in tenths of °C. (# NAs: `r ny_noaa %>% pull(tmax) %>% is.na() %>% sum()`) 

* **tmin**: Minimum temperature on a given day reported in tenths of °C. (# NAs: `r ny_noaa %>% pull(tmin) %>% is.na() %>% sum()`) 

The number of NA's in this dataset is concerning, especially if we want to do an analysis over the total period of time. However, for certain variables it is less significant than others. We can assume that for the days that the observations report NA values for prcp, snow, and snwd that it did not rain nor snow on those days. However, if we want to do an analysis of min and max temperature variations over the whole time period, we cannot feasibly accomplish this due to the missing values. 

* **Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. For snowfall, what are the most commonly observed values? Why?**

```{r clean_ny_noaa}

ny_noaa_df = ny_noaa %>% 
  separate(date, into = c("year", "month", "day"), sep = "-") %>% 
  mutate(
    tmax = as.integer(tmax),
    tmin = as.integer(tmin)
  ) %>% 
  mutate(
    prcp = prcp / 10,
    tmax = tmax / 10,
    tmin = tmin / 10
  )

ny_noaa_df %>% 
  count(snow) %>%
  drop_na() %>% 
  arrange(desc(n))
```

0mm is the most commonly reported value for snowfall, which makes sense as it only snows during the winter, and even then it is unlikely to snow during the winter. The following most common snowfall amounts range from ~ 5 - 100mm which is not much snowfall, probably what you would expect on a given day when it actually snows. 

* **Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?**

```{r plot_temp}

ny_noaa_df %>% 
  filter(month %in% c("01", "07")) %>% 
  group_by(id, month,year) %>% 
  summarise(mean_tmax = mean(tmax, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = mean_tmax, group = id)) +
  geom_point(alpha = .1) + 
  geom_path(alpha = .3) +
  facet_grid(. ~ month) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

It is tough to see with so many points and lines overlapping each other, but it is clear that all the weather stations in NY report similar temperature ranges between each other with the exception of some outliers. We can interpret from the plot that January is colder than July and that the temperature ranges from ~ -10°C to 10°C and 20°C to 35°C respectively.

* **Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.**

```{r plot_tmax_tmin}

tmax_tmin_plot = 
  ny_noaa_df %>% 
  ggplot(aes(x = tmax, y = tmin)) +
  geom_hex() +
  geom_abline(slope=1, intercept=0) +
  labs(
    title = "Minimum Temperature vs Maximum Temperature",
    x = "Max Temp (°C)", 
    y = "Minimum Temp (°C)",
    caption = "x = y added to plot"
  )
  
snow_plot = 
  ny_noaa_df %>% 
  drop_na(snow) %>% 
  filter(
    snow > 0, 
    snow < 100,
  ) %>% 
  ggplot(aes(x = snow, y = year)) +
  geom_density_ridges(scale = .85) +
  labs( 
    title = "Snowfall Data from NY Weather Stations: 1981 to 2010",
    x = "Snowfall Amount (mm)", 
    y = "Year"
  )

tmax_tmin_plot + snow_plot

```

